Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-05-09 paper.tex
text and figures updated

\documentclass[10pt,aps,pra,reprint,amsmath,amsfonts,amssymb,showpacs]{revtex4-1}
\usepackage{epsfig,psfrag,graphicx,bm,color}

@@

%NEW
\newcommand{\rmn}{\mathrm}
\newcommand{\bx}{{\bf x}}
\newcommand{\fg}{{F_\gamma}}
\newcommand{\sg}{{S_\gamma}}
\newcommand{\psf}{\theta_\rmn{res}}
\newcommand{\ph}{\rmn{ph}}
\newcommand{\eph}{E_\ph}
\newcommand{\nph}{N_\ph}
\newcommand{\vir}{\rmn{vir}}
\newcommand{\gal}{\rmn{gal}}
\newcommand{\sd}{\rmn{SD}}
\newcommand{\clu}{\rmn{cl}}
\newcommand{\sfe}{\rmn{sfe}}
\newcommand{\sub}{\rmn{sub}}
\newcommand{\msun}{M_\odot}
\newcommand{\ee}{E_\rmn{e}}
\newcommand{\stars}{\rmn{stars}}
\newcommand{\dust}{\rmn{dust}}
\newcommand{\lx}{L_\rmn{X}}
\newcommand{\s}{\rmn{s}}
\newcommand{\Kp}{\rmn{K}^\prime}
\newcommand{\Ip}{\rmn{I}^\prime}
\newcommand{\Js}{\rmn{J}^*}
\newcommand{\Jp}{\rmn{J}^\prime}
\newcommand{\B}{\rmn{B}}
\newcommand{\bsub}{\B_\rmn{sub}}
\newcommand{\qCR}{q_{\rmn{CR}-\ensuremath{\pi^0}}}
\newcommand{\ds}{{\sc DarkSUSY}}
\newcommand{\mlim}{M_\rmn{lim}}

%OLD
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\gev}{\rmn{GeV}}
\newcommand{\tev}{\rmn{TeV}}
\newcommand{\ev}{\rmn{eV}}
\newcommand{\dd}{\rmn{d}}
\newcommand{\mx}{\ensuremath{m_{\chi}}}
\newcommand{\ga}{\gamma}
\newcommand{\ngamma}{\ensuremath{N_{\gamma}}}
\newcommand{\ngammaj}{\ensuremath{N_{\gamma,j}}}
\newcommand{\sigmaannv}{\ensuremath{\langle\sigma v\rangle}}
\newcommand{\sigv}{\ensuremath{\sigma_v}}
\newcommand{\ic}{\rmn{IC}}
\newcommand{\egamma}{\ensuremath{E_{\gamma}}}
\newcommand{\dic}{\rmn{dIC}}
\newcommand{\fsr}{\rmn{FSR}}
\newcommand{\CR}{\rmn{CR}}
\newcommand{\rhos}{\ensuremath{\rho_s}}
\newcommand{\rs}{\ensuremath{r_s}}
\newcommand{\rvir}{r_{200}}
\newcommand{\mvir}{M_{200}}
\newcommand{\rhoc}{\ensuremath{\rho_c}}
\newcommand{\e}{\rmn{e}}
\newcommand{\eg}{E_\gamma}
\newcommand{\ir}{\rmn{ir}}

@@

Dark matter has been searched for in direct detection experiments
\cite{Pato:2010zk}, at accelerators
\cite{Ellis:2001hv,Baer:2006ff,Khachatryan:2011tk} and also in
indirect detection experiments looking for signals in the cosmic-ray
spectra of antiprotons, positrons, neutrinos and all of the
electromagnetic spectrum from radio waves to gamma rays
\cite{Bergstrom:2009ib}. So far, the improvements in direct detection
sensitivity have put this method into focus, but the situation may
change considerably the coming few years as the CERN LHC experiments
collect data, and new gamma-ray detectors are being planned, such as
the CTA \cite{Consortium:2010bc}. In fact, it has recently been
pointed out \cite{Bergstrom:2010gh} that a dedicated ground-based
gamma-ray detector would have potential that goes far beyond that of
the other methods, depending on presently unknown parameters in the
particle physics models for dark matter.

Among the astrophysical systems which will be very interesting to
detect, and study, with coming gamma-ray detectors (Fermi, HESS,
MAGIC, VERITAS, and eventually large detectors like CTA) belong galaxy
clusters. The most promising directions in which to search for a
gamma-ray annihilation signal (from the annihilation process itself,
and also the accompanying bremsstrahlung and inverse Compton
components coming from charged particles produced in the annihilations)
are basically three:

1. The galactic centre (g.c.). This is where all numerical simulations
of cold dark matter predict the highest density. However, the detailed
dark matter density in the very central part is difficult to predict,
due a to a possibly very complicated interplay between baryons, dark
matter, and the central galactic black hole. Also, it is a very
crowded region with many gamma-ray sources like pulsars and other
supernova remnants, which have to be subtracted from the data to
extract the dark matter induced signal. In fact, there is a recent
claim of an indication of a relatively light dark matter particle
contribution to the gamma-ray flux from the
g.c. \cite{2010arXiv1010.2752H}, but other hypotheses seem to work at
least as well \cite{2010arXiv1012.5839B}.

2. The dwarf spheroidal galaxies orbiting the Milky Way, like Segue-1,
Ursa Minor, Draco, Sagittarius, Sculptor, Carina or Willman-1
\cite{2009JCAP...01..016B,2010ApJ...720.1174A,2010JCAP...01..031S,2010JCAP...01..031S}. The
problem here is that the nature of many of these small, dark
matter-dominated galaxies is not entirely clear, and the velocity
dispersion estimates are based on rather small numbers of stars. Tidal
disruption and confusion with star clusters are other
complications. Thus the dark matter density profile is very uncertain
for most of them. Nonetheless, by stacking the data together from many
dwarf spheroidals these uncertainties can be made less severe, and
preliminary results from Fermi-LAT shows this method to give quite
promising results \cite{garde}.

3. Galaxy clusters. This possibility has been less studied, however we
noted in a previous Letter \cite{2009PhRvL.103r1302P} that there are
certain advantages that work in favour of this possible target for
gamma-ray detection of dark matter annihilation.  Galaxy clusters
constitute the most massive objects in our Universe that are forming
today. This causes their DM subhalo mass function to be less affected
by tidal stripping compared to galaxy sized halos that formed long
ago.  The annihilation luminosities of the DM halo component for
e.g. the Virgo cluster and the Draco dwarf scales in a way (see
\cite{2009PhRvL.103r1302P}) that the ratio of gamma-ray luminosities
from the smooth components is around 4, in favour of Virgo. In
addition, there may be a further enhancement due to substructure,
which to a large extent should be unaffected by tidal disruption, at
least in the outer regions.  According to a recent estimate
\cite{2011MNRAS.410.2309G}, more massive haloes tend to have a larger
mass fraction in subhalos.  For example, cluster size haloes typically
have 7.5 per cent of the mass within $r_{200}$ in substructures of
fractional mass larger than $10^{-5}$, which is 25 per cent higher
than galactic haloes \footnote{All halo masses and length scales in
  this paper are scaled to the currently favored value of Hubble's
  constant, $H_0 = 70\, \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define the
  virial mass $\mvir$ and virial radius $\rvir$ as the mass and radius
  of a sphere enclosing a mean density that is 200 times the critical
  density of the Universe $\rho_{\rmn{cr}}$.}.

@@

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to yield an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors.

For previous work related to dark matter in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J}.

%***********************************************
{\bf Anders, insert your flux table for various clusters, and discuss it}

This yields Fornax ($\mvir=10^{14}\,\msun$) and Virgo
($\mvir=2.1\times10^{14}\,\msun$) \del{\cite{}} as the prime targets for DM
observations, Perseus ($\mvir=7.7\times10^{14}\,\msun$) for CR induced emission
and we additionally decide in favor of the well studied cluster Coma
($\mvir=1.4\times10^{15}\,\msun$) for comparison. While studying these four
clusters in great detail, we complement the analysis by calculating the expected
gamma-ray fluxes of a nearby X-ray flux complete sample of galaxy clusters
(HIFLUGCS, \citealt{2002ApJ...567..716R})

@@

% --- section:  --- %
\section{Theory}
\label{sect:theory}
We start the section by discussing two different, but well motivated DM models;
leptophilic and supersymmetric DM. We then present the framework that is used to
calculate the gamma-ray emission for these DM models using an Einasto DM density
profile and the expected enhancement from DM substructures. This is followed by
an outline of the framework of inverse Compton emission where we take into
account photon fields of the cosmic microwave background (CMB), dust, and
starlight. We end by summarizing our formalism of calculating the CR-induced
pion decay gamma-ray emission which is thought to dominate the astrophysical
gamma-ray signal.

\subsection{Detecting Particle Dark Matter}
\label{sect:PF}
Besides the intrinsic interest in the gamma-ray flux from galaxy
clusters generated by conventional hadronic and electromagnetic
processes - which should be close to observability with the Fermi-LAT
data \cite{1997ApJ...487..529B,2007A&A...473...41E,2010MNRAS.409..449P}
- the possible contribution from WIMP dark matter is of greatest
interest. A WIMP (Weakly Interactive Massive Particle) which fulfils
the WMAP bounds on the relic density of cold dark matter (CDM) of
\cite{Komatsu:2010fb}
$$\Omega_{CDM}h^2=0.112\pm 0.0056,$$ will in many cases naturally give
a gamma-yield which may be observable. In addition, there are
possible enhancement effects known, such as the astrophysical boost
from dense substructure of dark matter halos, or the particle physics
boost from the Sommerfeld effect, which may increase the chances of
detection further.

There are three methods for detecting dark matter candidates which are
presently employed: First, at particle physics accelerators such as
CERN's LHC, one is now entering into a new energy regime which may
allow the production of the heavy, electrically neutral and long-lived
particles which may constitute dark matter. From these experiments one
may get the first glimpse of the mass scale beyond the standard model
where dark matter may reside. However, to really show that any of the
hypothetical, new particles created at the LHC is actually the dark
matter, one has to rely on the two other methods available for the
dark matter search, namely direct and indirect detection. Direct
detection methods, which are presently evolving rapidly use the feeble
interaction between dark matter particles of the galactic halo
and nuclei such as Germanium, Sodium, Iodine, Argon or Xenon to infer
the scattering cross section and a rough estimate of the mass of the
dark matter particle (for the currently most sensitive search, see
\cite{Aprile:2010um}). A characteristic of this method is that it
basically only depends on the local dark matter density (which is
rather well determined to be around $0.4$ GeV/cm$^3$) and the basic
cross section for dark matter - nucleus scattering. A disadvantage is
that it does benefit from the two enhancement mechanisms mentioned
above, boost from substructure and/or the Sommerfeld effect. Also, it
cannot be excluded - although it seems at present improbable - that
the local halo structure is such that the solar system happens to be
in an underdense region.

In indirect detection, on the other hand, in particular in the photon
channel, one searches for products of dark matter annihilation in the
galactic halo and beyond. Particularly interesting targets are the
galactic center (for a recent possible indication of a signal, see
\cite{Hooper:2010mq}), the dwarf galaxies surrounding the Milky Way
\cite{Strigari:2006rd,Essig:2009jx,2010JCAP...01..031S}, galaxy
clusters (the focus of this work)
\cite{Ghigna:1998vn,Lewis:2002mfa,Boyarsky:2006kc,2006A&A...455...21C,2009PhRvL.103r1302P},
and even the cosmological large scale structure
\cite{Bergstrom:2001jj,Ullio:2002pj,Taylor:2002zd,Elsaesser:2004ap,Cuoco:2010jb,Abazajian:2010sq,Abdo:2010dk,Zavala:2011tt}.

There is also a possibility to search for dark matter annihilation indirectly in
antimatter channels, like antiprotons or positrons. Here diffusion is, however,
very efficient which means that the directional signature which is so important
for gamma-rays, is completely erased. This method has, however, recently been
much in focus due to the surprising rise with energy of the positron-to-total
lepton ratio measured by the PAMELA satellite experiment \cite{Adriani:2008zr}
and the ``bump'' in the total electron plus positron spectrum detected by Fermi
\cite{Abdo:2009zk}. The viable particle physics models describing these results
contain some non-trivial, or even unnatural elements, however. First of all,
annihilation to quarks has to be very much suppressed, due to the
non-observation of an enhanced antiproton flux by PAMELA
\cite{Adriani:2010rc}. This means that only so-called {\em leptophilic} models,
where the annihilation only goes to leptons have to be considered. Here, one has
to have a sizeable fraction going to muons and tau leptons, as the shape of the
spectrum disfavours direct annihilation into electrons and positrons. To get
agreement with the measured absolute rates, one also has to have a large boost
factor (of the order of at least several hundred), which is difficult to achieve
with substructure alone. An additional enhancement of the annihilation rate has
to be invoked, for example given by the {\em Sommerfeld effect}. This basically
means that the usual connection between annihilation rate at decoupling (when
relative velocities $v/c$ were of order $0.2 - 0.3$), which gives the relic
density, and present-day annihilation in the halo (with velocities $v/c\sim
10^{-3}$) is broken. This would come about because the dark matter sector
contains additional light particles which mediates an interaction with almost
bound states which enhance the wave function at the relative origin, and
therefore the annihilation rate (for a pedagogical review of viable scenarios,
see \cite{ArkaniHamed:2008qn}). As a final difficulty of these models, the lack
of an inverse-Compton signature from the galactic center means that only shallow
density profiles can describe the galactic dark matter density distribution
\cite{Bertone:2008xr,Cirelli:2008pk,Bergstrom:2008ag}. (For a review of the dark
matter modelling of these effects, see \cite{Bergstrom:2009ib}. For a recent
treatment, showing still viable models, see \cite{Finkbeiner:2010sm}.)

A more standard explanation of the PAMELA/Fermi excess would be
pulsars or other supernova remnants (e.g.,
\cite{Hooper:2008kg,Ahlers:2009ae}). In that case, dark matter would
more likely be explained by the standard scenario of a WIMP, for
example the thoroughly studied lightest supersymmetric neutralino
(for reviews, see
\cite{Jungman:1995df,Bergstrom:2000pn,Bertone:2004pz}). We note that
the soon to be launched AMS-02 experiment on the International Space
Station \cite{ams02} may give interesting clues to the origin of the
antimatter excess.

>>>>

\documentclass[10pt,aps,pra,reprint,amsmath,amsfonts,amssymb,showpacs,nofootinbib,floatfix]{revtex4-1}
\usepackage{epsfig,psfrag,graphicx,bm,color}

@@

%NEW
\newcommand{\rmn}{\mathrm}
\newcommand{\bx}{{\bf x}}
\newcommand{\fg}{{F_\gamma}}
\newcommand{\sg}{{S_\gamma}}
\newcommand{\psf}{\theta_\rmn{res}}
\newcommand{\ph}{\rmn{ph}}
\newcommand{\eph}{E_\ph}
\newcommand{\nph}{N_\ph}
\newcommand{\vir}{\rmn{vir}}
\newcommand{\gal}{\rmn{gal}}
\newcommand{\sd}{\rmn{SD}}
\newcommand{\clu}{\rmn{cl}}
\newcommand{\sfe}{\rmn{sfe}}
\newcommand{\sub}{\rmn{sub}}
\newcommand{\msun}{M_\odot}
\newcommand{\ee}{E_\rmn{e}}
\newcommand{\stars}{\rmn{stars}}
\newcommand{\dust}{\rmn{dust}}
\newcommand{\lx}{L_\rmn{X}}
\newcommand{\s}{\rmn{s}}
\newcommand{\Kp}{\rmn{K}^\prime}
\newcommand{\Ip}{\rmn{I}^\prime}
\newcommand{\Js}{\rmn{J}^*}
\newcommand{\Jp}{\rmn{J}^\prime}
\newcommand{\B}{\rmn{B}}
\newcommand{\bsub}{\B_\rmn{sub}}
\newcommand{\qCR}{q_{\rmn{CR}-\ensuremath{\pi^0}}}
\newcommand{\ds}{{\sc DarkSUSY}}
\newcommand{\mlim}{M_\rmn{lim}}
\newcommand{\sm}{\rmn{sm}}
\newcommand{\mpc}{\rmn{Mpc}}
\newcommand{\kpc}{\rmn{kpc}}
\newcommand{\mev}{\rmn{MeV}}
\newcommand{\cm}{\rmn{cm}}
%OLD
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\gev}{\rmn{GeV}}
\newcommand{\tev}{\rmn{TeV}}
\newcommand{\ev}{\rmn{eV}}
\newcommand{\dd}{\rmn{d}}
\newcommand{\mx}{\ensuremath{m_{\chi}}}
\newcommand{\ga}{\gamma}
\newcommand{\ngamma}{\ensuremath{N_{\gamma}}}
\newcommand{\ngammaj}{\ensuremath{N_{\gamma,j}}}
\newcommand{\sigmaannv}{\ensuremath{\langle\sigma v\rangle}}
\newcommand{\sigv}{\ensuremath{\sigma_v}}
\newcommand{\ic}{\rmn{IC}}
\newcommand{\egamma}{\ensuremath{E_{\gamma}}}
\newcommand{\dic}{\rmn{dIC}}
\newcommand{\fsr}{\rmn{FSR}}
\newcommand{\CR}{\rmn{CR}}
\newcommand{\rhos}{\ensuremath{\rho_s}}
\newcommand{\rs}{\ensuremath{r_s}}
\newcommand{\rvir}{r_{200}}
\newcommand{\mvir}{M_{200}}
\newcommand{\rhoc}{\ensuremath{\rho_c}}
\newcommand{\e}{\rmn{e}}
\newcommand{\eg}{E_\gamma}
\newcommand{\ir}{\rmn{ir}}

@@

Dark matter (DM) has been searched for in direct detection experiments
\cite{Pato:2010zk}, at accelerators
\cite{Ellis:2001hv,Baer:2006ff,Khachatryan:2011tk} and also in
indirect detection experiments looking for signals in the cosmic-ray
spectra of antiprotons, positrons, neutrinos and all of the
electromagnetic spectrum from radio waves to gamma rays
\cite{Bergstrom:2009ib}. So far, the improvements in direct detection
sensitivity have put this method into focus, but the situation may
change considerably the coming few years as the CERN LHC experiments
collect data, and new gamma-ray detectors are being planned, such as
the CTA \cite{Consortium:2010bc}. In fact, it has recently been
pointed out \cite{Bergstrom:2010gh} that a dedicated ground-based
gamma-ray detector would have potential that goes far beyond that of
the other methods, depending on presently unknown parameters in the
particle physics models for DM.

Among the astrophysical systems which will be very interesting to
detect, and study, with coming gamma-ray detectors (Fermi, HESS,
MAGIC, VERITAS, and eventually large detectors like CTA) belong galaxy
clusters. The most promising directions in which to search for a
gamma-ray annihilation signal (from the annihilation process itself,
and also the accompanying bremsstrahlung and inverse Compton (IC)
components coming from charged particles produced in the annihilations)
are basically three:

1. The galactic center (g.c.). This is where all numerical simulations
of cold dark matter (CDM) predict the highest density. However, the
detailed DM density in the very central part is difficult to predict,
due a to a possibly very complicated interplay between baryons, DM,
and the central galactic black hole. Also, it is a very crowded region
with many gamma-ray sources like pulsars and other supernova remnants,
which have to be subtracted from the data to extract the DM induced
signal. In fact, there is a recent claim of an indication of a
relatively light DM particle contribution to the gamma-ray flux from
the g.c. \cite{2010arXiv1010.2752H}, but other hypotheses seem to work
at least as well \cite{2010arXiv1012.5839B}.

2. The dwarf spheroidal galaxies orbiting the Milky Way, like Segue-1,
Ursa Minor, Draco, Sagittarius, Sculptor, Carina or Willman-1
\cite{2009JCAP...01..016B,2010ApJ...720.1174A,2010JCAP...01..031S,2010JCAP...01..031S}. The
problem here is that the nature of many of these small, dark
matter-dominated galaxies is not entirely clear, and the velocity
dispersion estimates are based on rather small numbers of stars. Tidal
disruption and confusion with star clusters are other
complications. Thus the DM density profile is very uncertain for most
of them. Nonetheless, by stacking the data together from many dwarf
spheroidals these uncertainties can be made less severe, and
preliminary results from Fermi-LAT shows this method to give quite
promising results \cite{garde}.

3. Galaxy clusters. This possibility has been less studied, however we
noted in a previous Letter \cite{2009PhRvL.103r1302P} that there are
certain advantages that work in Cavour of this possible target for
gamma-ray detection of DM annihilation.  Galaxy clusters constitute
the most massive objects in our Universe that are forming today. This
causes their DM subhalo mass function to be less affected by tidal
stripping compared to galaxy sized halos that formed long ago.  The
annihilation luminosities of the DM halo component for e.g. the Virgo
cluster and the Draco dwarf scales in a way (see
\cite{2009PhRvL.103r1302P}) that the ratio of gamma-ray luminosities
from the smooth components is around 4, in Cavour of Virgo. In
addition, there may be a further enhancement due to substructure,
which to a large extent should be unaffected by tidal disruption, at
least in the outer regions.  According to a recent estimate
\cite{2011MNRAS.410.2309G}, more massive haloes tend to have a larger
mass fraction in subhalos.  For example, cluster size haloes typically
have 7.5 per cent of the mass within $r_{200}$ in substructures of
fractional mass larger than $10^{-5}$, which is 25 per cent higher
than galactic haloes \footnote{All halo masses and length scales in
  this paper are scaled to the currently favored value of Hubble's
  constant, $H_0 = 70\, \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define
  the virial mass $\mvir$ and virial radius $\rvir$ as the mass and
  radius of a sphere enclosing a mean density that is 200 times the
  critical density of the Universe $\rho_{\rmn{cr}}$.}.

@@

{\bf Anders, insert your flux table for various clusters, and discuss it, DONE}

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to yield an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. In Table~\ref{tab:flux_tab} we show
the gamma-ray flux from four clusters, which are all targets with a
high gamma-ray yield. We derive the emission from two different DM
models and a hadronic CR model and find that Virgo has the highest
gamma-ray flux which emerge from annihilating DM. If the boost from
Sommerfeld enhancement is small, however, then the gamma-ray emission
induced by CRs is dominating most clusters with the exception of a few
clusters where Fornax had the highest flux from DM. In particular, we
find Fornax ($\mvir=10^{14}\,\msun$) and Virgo
($\mvir=2.1\times10^{14}\,\msun$) \del{\cite{}} as the prime targets
for DM observations, Perseus ($\mvir=7.7\times10^{14}\,\msun$) for CR
induced emission and we additionally decide in favor of the well
studied cluster Coma ($\mvir=1.4\times10^{15}\,\msun$) for
comparison. While studying these four clusters in great detail, we
complement the analysis by calculating the expected gamma-ray fluxes
of a nearby X-ray flux complete sample of galaxy clusters (HIFLUGCS,
\citealt{2002ApJ...567..716R}). For previous work related to DM in
clusters, see, e.g., \cite{2006A&A...455...21C,2009PhRvD..80b3005J}.

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters.}
\begin{tabular}{c  c c c  c  c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} \\
         & DM-LP $^{(1)}$ & DM-BM-$\Kp$ $^{(2)}$ & CR-$\pi^0$ $^{(3)}$ 
         & & DM-LP $^{(1)}$ & DM-BM-$\Kp$ $^{(2)}$ & CR-$\pi^0$ $^{(3)}$ \\
 \hline
Fornax               & $4.8\times10^{-8}$ & $1.3\times10^{-10}$ &  $2.2\times10^{-10}$ 
& \,\,\,\,\, & $5.6\times10^{-12}$  & $3.6\times10^{-13}$ & $8.3\times10^{-14}$ \\
Coma                 & $1.3\times10^{-8}$ & $8.3\times10^{-11}$ &  $4.0\times10^{-9}$  
& \,\,\,\,\, & $1.4\times10^{-12}$  & $2.3\times10^{-13}$ & $1.5\times10^{-12}$ \\
Virgo                & $1.2\times10^{-7}$ & $4.0\times10^{-10}$ &  $8.7\times10^{-9}$  
& \,\,\,\,\, & $1.3\times10^{-11}$  & $1.1\times10^{-12}$ & $3.2\times10^{-12}$ \\
Perseus \,\,\,\,\,\, & $1.4\times10^{-8}$ & $7.1\times10^{-11}$ &  $1.4\times10^{-8}$  
& \,\,\,\,\, & $1.4\times10^{-12}$  & $2.0\times10^{-13}$ & $5.2\times10^{-12}$ \\
\hline
\hline
\end{tabular}
\begin{quote}
 Notes: \\ 
 (1) The total gamma-ray flux from leptophilic (LP) DM where both the boost 
 from substructures and Sommerfeld enhancement are included.\\
 (2) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
 model where the boost from substructures is included.\\
 (3) Gamma-ray flux induced by CR protons.
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

@@

% --- section:  --- %
\section{Theory}
\label{sect:theory}
We start the section by discussing two different, but well motivated
DM models; leptophilic and supersymmetric DM. We then present the
framework that is used to calculate the gamma-ray emission for these
DM models using an Einasto DM density profile and the expected
enhancement from DM substructures. This is followed by an outline of
the framework of IC emission where we take into account photon fields
of the cosmic microwave background (CMB), dust, and starlight. We end
by summarizing our formalism of calculating the CR-induced pion decay
gamma-ray emission which is thought to dominate the astrophysical
gamma-ray signal.

\subsection{Detecting Particle Dark Matter}
\label{sect:PF}
Besides the intrinsic interest in the gamma-ray flux from galaxy
clusters generated by conventional hadronic and electromagnetic
processes - which should be close to observability with the Fermi-LAT
data
\cite{1997ApJ...487..529B,2007A&A...473...41E,2010MNRAS.409..449P} -
the possible contribution from WIMP DM is of greatest interest. A WIMP
(Weakly Interactive Massive Particle) which fulfils the WMAP bounds on
the relic density of CDM of \cite{Komatsu:2010fb}
$$\Omega_{CDM}h^2=0.112\pm 0.0056,$$ will in many cases naturally give
a gamma-yield which may be observable. In addition, there are possible
enhancement effects known, such as the astrophysical boost from dense
substructure of DM halos, or the particle physics boost from the
Sommerfeld effect, which may increase the chances of detection
further.

There are three methods for detecting DM candidates which are
presently employed: First, at particle physics accelerators such as
CERN's LHC, one is now entering into a new energy regime which may
allow the production of the heavy, electrically neutral and long-lived
particles which may constitute DM. From these experiments one may get
the first glimpse of the mass scale beyond the standard model where DM
may reside. However, to really show that any of the hypothetical, new
particles created at the LHC is actually the DM, one has to rely on
the two other methods available for the DM search, namely direct and
indirect detection. Direct detection methods, which are presently
evolving rapidly use the feeble interaction between DM particles of
the galactic halo and nuclei such as Germanium, Sodium, Iodine, Argon
or Xenon to infer the scattering cross section and a rough estimate of
the mass of the DM particle (for the currently most sensitive search,
see \cite{Aprile:2010um}). A characteristic of this method is that it
basically only depends on the local DM density (which is rather well
determined to be around $0.4$ GeV/cm$^3$) and the basic cross section
for DM - nucleus scattering. A disadvantage is that it does benefit
from the two enhancement mechanisms mentioned above, boost from
substructure and/or the Sommerfeld effect. Also, it cannot be excluded
- although it seems at present improbable - that the local halo
structure is such that the solar system happens to be in an underdense
region.

In indirect detection, on the other hand, in particular in the photon
channel, one searches for products of DM annihilation in the galactic
halo and beyond. Particularly interesting targets are the galactic
center (for a recent possible indication of a signal, see
\cite{Hooper:2010mq}), the dwarf galaxies surrounding the Milky Way
\cite{Strigari:2006rd,Essig:2009jx,2010JCAP...01..031S}, galaxy
clusters (the focus of this work)
\cite{Ghigna:1998vn,Lewis:2002mfa,Boyarsky:2006kc,2006A&A...455...21C,2009PhRvL.103r1302P},
and even the cosmological large scale structure
\cite{Bergstrom:2001jj,Ullio:2002pj,Taylor:2002zd,Elsaesser:2004ap,Cuoco:2010jb,Abazajian:2010sq,Abdo:2010dk,Zavala:2011tt}.

There is also a possibility to search for DM annihilation indirectly
in antimatter channels, like antiprotons or positrons. Here diffusion
is, however, very efficient which means that the directional signature
which is so important for gamma-rays, is completely erased. This
method has, however, recently been much in focus due to the surprising
rise with energy of the positron-to-total lepton ratio measured by the
PAMELA satellite experiment \cite{Adriani:2008zr} and the ``bump'' in
the total electron plus positron spectrum detected by Fermi
\cite{Abdo:2009zk}. The viable particle physics models describing
these results contain some non-trivial, or even unnatural elements,
however. First of all, annihilation to quarks has to be very much
suppressed, due to the non-observation of an enhanced antiproton flux
by PAMELA \cite{Adriani:2010rc}. This means that only so-called {\em
  leptophilic} models, where the annihilation only goes to leptons
have to be considered. Here, one has to have a sizeable fraction going
to muons and tau leptons, as the shape of the spectrum disfavors
direct annihilation into electrons and positrons. To get agreement
with the measured absolute rates, one also has to have a large boost
factor (of the order of at least several hundred), which is difficult
to achieve with substructure alone. An additional enhancement of the
annihilation rate has to be invoked, for example given by the {\em
  Sommerfeld effect}. This basically means that the usual connection
between annihilation rate at decoupling (when relative velocities
$v/c$ were of order $0.2 - 0.3$), which gives the relic density, and
present-day annihilation in the halo (with velocities $v/c\sim
10^{-3}$) is broken. This would come about because the DM sector
contains additional light particles which mediates an interaction with
almost bound states which enhance the wave function at the relative
origin, and therefore the annihilation rate (for a pedagogical review
of viable scenarios, see \cite{ArkaniHamed:2008qn}). As a final
difficulty of these models, the lack of an IC signature from the
galactic center means that only shallow density profiles can describe
the galactic DM density distribution
\cite{Bertone:2008xr,Cirelli:2008pk,Bergstrom:2008ag}. (For a review
of the DM modelling of these effects, see \cite{Bergstrom:2009ib}. For
a recent treatment, showing still viable models, see
\cite{Finkbeiner:2010sm}.)

A more standard explanation of the PAMELA/Fermi excess would be
pulsars or other supernova remnants (e.g.,
\cite{Hooper:2008kg,Ahlers:2009ae}). In that case, DM would more
likely be explained by the standard scenario of a WIMP, for example
the thoroughly studied lightest supersymmetric neutralino (for
reviews, see
\cite{Jungman:1995df,Bergstrom:2000pn,Bertone:2004pz}). We note that
the soon to be launched AMS-02 experiment on the International Space
Station \cite{ams02} may give interesting clues to the origin of the
antimatter excess.

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) has caused intense
activity in the theoretical DM community. At first, there was
skepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
DM halo substructure has been realized for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieve the boost factor of the order of a few hundred
to a thousand in the solar neighborhood, as substructure according to
numerical simulations is most important in the outer portions of the
halo, due to tidal stripping of subhalos in the inner part. Another
potentially very important effect which may explain the large boost
had been found a few years earlier, namely the Sommerfeld
enhancement. This effect, was computed for electromagnetism by Arnold
Sommerfeld many years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of DM
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probability for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
$$
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=
\left|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}\right|,
$$ 
with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the relative
velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small velocities. For a
Yukawa-like particle of mass $m_\phi$, mediating a weak attractive force with
coupling constant $\alpha_Y$ between DM particles $\chi$ with mass
$m_\chi$, the small-velocity limit of the enhancement becomes instead
$$
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
$$ In the general case, with mediators that may also excite virtual
charged particles in the DM sector, one has to solve numerically a
coupled system of differential equations using appropriate boundary
conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depending however on the exact parameters
of the theory. The effect is usually strongly velocity-dependent,
depending on velocity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such as a galaxy cluster) with large velocity dispersion
$v_{\rm cl}$ the Sommerfeld enhancement will be smaller than the one
expected in a single galaxy such as the Milky Way, roughly by a factor
$v_{\rm MW}/v_{\rm cl}$.

As leptophilic models, apart from being slightly contrived from the
particle physics point of view, are also rather limited by several
sets of astrophysical data, we take advantage of the recent reanalysis
\cite{Finkbeiner:2010sm} to define a benchmark leptophilic model that
is still viable.  It is found that the bounds from WMAP5 approximately
imply a Sommerfeld boost factor $S$ which has to satisfy $S(v\to
0)\lesssim 200/f\ (m_Y/1$ GeV$)$, where $f$ is the fraction of energy
from DM annihilation that ionizes the intergalactic medium,
$f\sim 0.7$ for annihilation to electrons and a factor of a few smaller
for all other standard model final states.

Taking into account the whole cosmic history of leptophilic models,
also utilizing limit on spectral and polarization distortions of the
CMB, the maximal value for the boost factor for a 1 to 2 TeV particle
and sub-GeV force carriers is found to be \cite{Finkbeiner:2010sm}
between 400 and 800. Although lower than the first estimates (e.g.,
\cite{Bergstrom:2009fa,Meade:2009iu}) this is still enough to explain
the PAMELA and Fermi excess, given other astrophysical uncertainties.
